Lesson 4
========================================================

***
getwd()
### Scatterplots and Perceived Audience Size
Notes:
```{r}
getwd()
list.files()

pf = read.csv('pseudo_facebook.tsv', sep='\t')
library(ggplot2)
qplot(x = friend_count, y = friendships_initiated, data=pf)
qplot(x = age, y = friend_count, data=pf)
```

***

### Scatterplots
Notes:

```{r Scatterplots}
ggplot(aes(x=tenure/365, y=likes), data=pf) + geom_point() + xlim(1,3) + xlab('Years on Network')
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_point(alpha=1/20) + xlab('Age of User')+
  xlim(12, 90)

ggplot(aes(x=tenure/365, y = likes_received), data=pf) + geom_jitter(alpha=1/5) + xlab('Number of Years on Network') + ylab('Likes Received') 

```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_jitter(alpha=1/20) + xlim(12,90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
?coord_trans
ggplot(pf, aes(x=age, y=friend_count)) + geom_point(alpha=1/20) + coord_trans( y="sqrt") +
  xlim(12,90)

ggplot(pf, aes(x=age, y=friend_count)) + geom_point(alpha=1/20, position = position_jitter(h= 0)) + coord_trans( y="sqrt") +  xlim(12,90)
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x=age, y=friendships_initiated), data=pf) + geom_jitter(alpha=1/10, position=position_jitter(h=0)) + coord_trans(y='sqrt') + 
  xlim(12,90) 


```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
install.packages('dplyr', dependencies = TRUE)
library(dplyr)

age_groups <- group_by(pf, age)
pf.fc_byage <- summarize(age_groups,
                         fc_mean = mean(friend_count),
                         fc_median = median(as.numeric(friend_count)),
                         n = n())

pf.fc_byage <- arrange(pf.fc_byage, age)
head(pf.fc_byage, 5)

```

Create your plot!

```{r Conditional Means Plot}

ggplot(aes(x=age, y=fc_mean), data=pf.fc_byage) + geom_line(lwd=2, color='yellow') + xlim(12,90) + xlab('Age') + ylab('Mean Friend Count') + ggtitle('Mean Friend Count vs Age')
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}

```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}

```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(                 , cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}

```

***

### Strong Correlations
Notes:

```{r Strong Correlations}

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}

```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
```

Create your plot!

```{r Temp vs Month}

```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}

```

***

### Age with Months Means

```{r Age with Months Means}

```

Programming Assignment
```{r Programming Assignment}

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

