Lesson 4
========================================================

***
getwd()
### Scatterplots and Perceived Audience Size
Notes:
```{r}
getwd()
list.files()

pf = read.csv('pseudo_facebook.tsv', sep='\t')
library(ggplot2)
qplot(x = friend_count, y = friendships_initiated, data=pf)
qplot(x = age, y = friend_count, data=pf)
```

***

### Scatterplots
Notes:

```{r Scatterplots}
ggplot(aes(x=tenure/365, y=likes), data=pf) + geom_point() + xlim(1,3) + xlab('Years on Network')
```

***

#### What are some things that you notice right away?
Response:

***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_point(alpha=1/20) + xlab('Age of User')+
  xlim(12, 90)

ggplot(aes(x=tenure/365, y = likes_received), data=pf) + geom_jitter(alpha=1/5) + xlab('Number of Years on Network') + ylab('Likes Received') 

```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_jitter(alpha=1/20) + xlim(12,90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:

```{r Coord_trans()}
?coord_trans
ggplot(pf, aes(x=age, y=friend_count)) + geom_point(alpha=1/20) + coord_trans( y="sqrt") +
  xlim(12,90)

ggplot(pf, aes(x=age, y=friend_count)) + geom_point(alpha=1/20, position = position_jitter(h= 0)) + coord_trans( y="sqrt") +  xlim(12,90)
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}

```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x=age, y=friendships_initiated), data=pf) + geom_jitter(alpha=1/10, position=position_jitter(h=0)) + coord_trans(y='sqrt') + 
  xlim(12,90) 


```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:

```{r Conditional Means}
install.packages('dplyr', dependencies = TRUE)
library(dplyr)

age_groups <- group_by(pf, age)
pf.fc_byage <- summarize(age_groups,
                         fc_mean = mean(friend_count),
                         fc_median = median(as.numeric(friend_count)),
                         n = n())

pf.fc_byage <- arrange(pf.fc_byage, age)
head(pf.fc_byage, 5)

```

Create your plot!

```{r Conditional Means Plot}

ggplot(aes(x=age, y=fc_mean), data=pf.fc_byage) + geom_line(lwd=2, color='blue') + xlim(12,90) + xlab('Age') + ylab('Mean Friend Count') + ggtitle('Mean Friend Count vs Age')
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x=age, y=friend_count), data=pf) + geom_jitter(alpha=1/20, color='orange') + 
  geom_line(stat='summary', fun.y=mean, lwd=2, col='red') + 
  geom_line(stat='summary', fun.y=quantile, fun.args = list(probs=0.1), linetype = 2) + 
  geom_line(stat='summary', fun.y=quantile, fun.args=list(probs=0.9), linetype=4) + 
  coord_cartesian(ylim=c(0,1500), xlim=c(12,90))
```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
?cor.test
cor.test(x=pf$age, y=pf$friend_count, method="pearson")
with(pf, cor.test(x=age, y=friend_count, method="pearson"))
with(subset(pf, age<70), cor.test(x=age, y=friend_count, method="pearson"))

```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(            subset(pf, age<=70)     , cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(www_likes_received, likes_received), data=pf) + geom_point() + 
  coord_cartesian(xlim=c(0,10000), ylim=c(0,10000)) + 
  geom_smooth(method="lm")

with(pf, cor.test(www_likes_received, likes_received, method="pearson"))

with(pf, lm(www_likes_received ~ likes_received))


```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
install.packages('alr3', dependencies=TRUE)
library(alr3)
data(Mitchell)
?Mitchell
soildf = Mitchell
ggplot(aes(x=Month, y=Temp), data = soildf) + geom_point()
with(soildf, cor.test(Month, Temp, method = "pearson"))
ggplot(aes(x=Month, y=Temp), data= (subset(soildf , Month<12))) + geom_point()

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(soildf, cor.test(x=Month, y=Temp, method="pearson"))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
install.packages('alr3')
library(alr3)
```

Create your plot!

```{r Temp vs Month}
years = c('One','Two', 'Three' , 'Four', 'Five', 'Six', 'Seven', 'Eight',
          'Nine', 'Ten', 'Eleven', 'Twelve', 'Thirteen', 'Fourteen','Fifteen',
          'Sixteen', 'Seventeen')
j <- 1
for (i in seq(0,204,12)){
  assign(years[j], soildf[(i+1):(i+12),])
  j <- j + 1
}

library(ggplot2)
ggplot(aes(x=Month, y=Temp), data=Six) + geom_point()

ggplot(aes(x=Month, y=Temp), data=soildf) + geom_point() + scale_x_continuous(breaks=seq(0,203,12))

ggplot(aes(x=Month%%12, y=Temp), data=soildf) + geom_point() + 
  scale_x_continuous(breaks=seq(0,11,1)) + xlab('Month') + ggtitle(label='Soil Temperature vs Month')

```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}

```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
pf$age_by_month <- pf$age + (12-pf$dob_month)/12
```

***

### Age with Months Means

```{r Age with Months Means}
library(dplyr)
age_groups_month <- group_by(pf, age_by_month)
pf.fc_by_age_months <- summarise(age_groups_month,
                        friend_count_mean = mean(friend_count),
                        friend_count_median = as.numeric(median(friend_count)),
                        n = n())

pf.fc_by_age_months <- arrange(pf.fc_by_age_months, age_by_month)
```

Programming Assignment
```{r Programming Assignment}
ggplot(aes(age_by_month), data = pf.fc_by_age_months) + 
  geom_line(aes(y =friend_count_mean,  color='Mean')) + 
  geom_line(aes(y = friend_count_median,  color='Median')) + coord_cartesian(xlim=(c(12,71)), ylim=c(0,500)) + 
  labs(x= 'Age of User', y='Friend Count', color='')

?legend
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}

```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

