Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
pf = read.csv('pseudo_facebook.tsv', sep='\t')
library(ggplot2)
library(dplyr)
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_histogram()
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}
gender_groups <- group_by(pf, gender, age)
pf_fc_by_gender <- summarise(gender_groups, 
                             mean_fc = mean(friend_count),
                             median_fc = median(friend_count),
                             n = n ())
ggplot(aes(x=age, y= median_fc, color='Male'), data = subset(pf_fc_by_gender, gender == 'male')) + geom_line(lwd=1.2) + geom_line(aes(x = age, y=median_fc, color = 'Female'), data = subset(pf_fc_by_gender, gender =='female'), lwd=1.2) + coord_cartesian(xlim=c(12,90)) + labs(y = 'Median Friend Count', title = 'Median Friend Count vs Age by Gender') + theme(legend.title=element_blank(), plot.title=element_text(hjust=0.5)) + scale_fill_discrete(name='Gender')


# Chain commands using %>%
pf_fc_by_gender <- pf_fc_by_gender %>% 
  filter(!is.na(gender)) %>%
  arrange(age)

ggplot(aes(x=age, y = median_fc), data = pf_fc_by_gender) + geom_line(aes(color=gender)) + coord_cartesian(xlim=c(12,90)) + labs(y='Median Friend Count', title = 'Median Friend Count vs Age by Gender') + 
  theme(plot.title=element_text(hjust=0.5))

```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)

pf_fc_by_gender <- dcast(pf_fc_by_gender, age ~ gender, value.var = 'median_fc')
```


***

### Ratio Plot
Notes:

```{r Ratio Plot}
ggplot(aes(x=age, y=female/male), data = pf_fc_by_gender) + geom_line(lwd = 1.2, color='navy') + 
  coord_cartesian(xlim=c(12,90)) + labs(y='Female/Male FC Ratio' , title = 'Female/Male FC Ratio vs Age of User') + 
  theme(plot.title=element_text(hjust=0.5)) + scale_x_continuous(breaks=seq(12,90,8)) + 
  scale_y_continuous(breaks=seq(0.5,2.5,0.5)) + geom_hline(aes(yintercept=1), lwd = 0.8, linetype = 3, color='red')
```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
pf$year_joined <- with(pf, floor(2014 - tenure/365))
```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
pf$year_joined.bucket <- cut(pf$year_joined, breaks = c(2004, 2009, 2011, 2012, 2014))
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
age_bucket <- group_by(pf, age, year_joined.bucket)
pf_by_age_bucket <- summarise(age_bucket,
                              median_fc = median(friend_count),
                              mean_fc = mean(friend_count),
                              n = n())

pf_by_age_bucket <- filter(pf_by_age_bucket, !is.na(year_joined.bucket))

ggplot(aes(x = age, y = median_fc), data = pf_by_age_bucket) + geom_line(aes(color=year_joined.bucket), lwd=1.2) + coord_cartesian(xlim=c(12,90)) + labs(title='Median Friend Count vs Age by Year Joined') + scale_color_discrete(name='Year Joined Bucket')

ggplot(aes(x=age, y=friend_count), data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color= year_joined.bucket), stat = 'summary', fun.y = median) + scale_color_discrete(name = 'Year Joined Bucket')


```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
ggplot(aes(x=age, y=friend_count), data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color= year_joined.bucket), stat = 'summary', fun.y = mean) + scale_color_discrete(name = 'Year Joined Bucket') + geom_line( stat = 'summary', fun.y=mean, linetype = 2, lwd = 1.2) + 
  scale_x_continuous(breaks=seq(12,90,8)) + coord_cartesian(xlim=c(12,90))

```

***

### Friending Rate
Notes:

```{r Friending Rate}
with(subset(pf, tenure > 1), summary(friend_count/tenure))

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
ggplot(aes(x=tenure, y = friendships_initiated / tenure), data = subset(pf, tenure > 1)) + geom_smooth(aes(color=year_joined.bucket), lwd = 0.8)
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo = read.csv('yogurt.csv', sep=",")
str(yo)

yo$id <- factor(yo$id)
str(yo)

qplot(x=price, data=yo, binwidth = 2, color = I('black'), fill = I('blue'))
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
?transform
yo <- transform(yo, all.purchases = strawberry + pina.colada + blueberry + plain + mixed.berry)
qplot(x = all.purchases, data = yo, binwidth = 1, color=I('black'), fill = I('green'))

```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(aes(x=time, y = price), data=yo) + geom_point(alpha = 1/10, color = 'red') + labs(y ='Price (cents)')
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(4230)
sample.ids <- sample(levels(yo$id), 16)
ggplot(aes(x = time, y = price), data = subset(yo, id %in% sample.ids)) + facet_wrap(~id) + 
  geom_line() + geom_point(aes(size=all.purchases), pch = 1) 

# Purchases over time sized by price of yogurt. I wonder if multiple types of yogurt were purchased
# if the price is the average price per yogurt. This graph does not provide much information. 
# I would expect houses to increase purchases when the price is lower. That does not seem to be 
# the case with this sample of households. The second set of plots is number of units purchased 
# versus price for the sample of households. 

set.seed(1204)
sample.ids <- sample(levels(yo$id), 8)
ggplot(aes(x = time, y= all.purchases), data = subset(yo, id %in% sample.ids)) + facet_wrap(~id) + 
  geom_line(color='blue') + geom_point(aes(size = price), color='green')

ggplot(aes(x = price, y= all.purchases), data = subset(yo, id %in% sample.ids)) + facet_wrap(~id) + 
  geom_line(color='blue') + geom_point(color='green', size = 1.8)

ggplot(aes(x = price, y=all.purchases), data = yo) + geom_point(stat = 'summary' , fun.y = mean)

```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:
```{r}

install.packages('GGally')
library(GGally)
?nasa
data(nasa)
df <- nasa
theme_set(theme_minimal)

# Graphing all correlations at once
set.seed(1836)
pf_subset <- pf[, 2:10]
pf_subset <- pf_subset[sample.int(nrow(pf_subset), 1000),]
ggpairs(pf_subset, columns = 1:8)
str(pf_subset)
?ggpairs
?sample.int

```
***

### Scatterplot Matrix
Notes:

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
library(reshape2)
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

library(ggplot2)
ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(name = 'Value', colours = colorRampPalette(c("green", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

